---
title: "Coding workshop: Week 4"
name: "your name here"
date: 2023-04-26
format: docx
---

# 0. set up

```{r message = FALSE}

```

# 1. Data

What is the first step to working with data?

```{r}

```

Are there any missing observations?  

Use `install.packages("naniar")` in the console before running the chunk below.

```{r message = FALSE, warning = FALSE}

```

# 2. Do sugar maple stem lengths differ between reference and calcium-treated watersheds in 2003?

## a. initial wrangling

```{r}

```

## b. summary statistics

Remember, we're interested in **stem lengths** in **2003** between **reference** and **calcium-treated** watersheds. What groups would be useful if that was the case?

```{r}
lengths_2003_summary <- maples_2003 %>% 
  group_by(watershed) %>% 
  summarize(mean_l = mean(stem_length),
            sd_l = sd(stem_length),
            var_l = var(stem_length),
            count_l = length(stem_length),
            se_l = sd_l/sqrt(count_l),
            margin_l = qt(0.95, df = count_l - 1) * se_l) 

lengths_2003_summary

# not getting the digits after the decimal point that you're expecting?
# try `as.data.frame()` piped in after the summarize call.
```

## c. check for normally distributed data

```{r}

```

## c. Check for equal variances

With an F-test, you can ask: are the sample variances between my two groups equal?  
The assumption is that your data are normally distributed.

$$
H0: s^2_1 = s^2_2 \\
H1: s^2_1 \neq s^2_2
$$

```{r}

```

## c. decide on a critical value

Two-tailed with significance level $\alpha$ = 0.05:

```{r}

```

If your test statistic is less than -1.98 or greater than 1.98, then you have evidence to reject the null hypothesis.

## d. Do a t-test 

```{r}

```

## e. Calculate Cohen's d effect size

**Cohen's d** is a measure of how many standard deviations apart the two sample means are. 

$$
Cohen's d = \frac{\bar{x_1} - \bar{x_2}}{\sqrt{(s^2_1 + s^2_2)/2}}
$$

Note that you are using sample means in the numerator and sample variances in the denominator.  

We can calculate this by hand (use `install.packages("data.table")` in the console before running the chunk below):  

```{r message = FALSE}
library(data.table)

# create a data frame in data table format from lengths_2003_summary
lengths_dt <- setDT(lengths_2003_summary)

# pull out mean and variance values from the data table 
mean_ref_2003 <- lengths_dt[watershed == "Reference", mean_l]
mean_w1_2003 <- lengths_dt[watershed == "Calcium-treated", mean_l]
var_ref_2003 <- lengths_dt[watershed == "Reference", var_l]
var_w1_2003 <- lengths_dt[watershed == "Calcium-treated", var_l]

# calculate Cohen's d

```

Or using a function in a package:  

Use `install.packages("effsize")` in the console before running the chunk below.

```{r}

```

Compare the two calculations:

```{r}

```

## make a plot

```{r fig.width = 8, fig.height = 6}
ggplot(data = lengths_2003_summary, aes(x = watershed, y = mean_l, color = watershed)) +
  geom_point(size = 3) +
  geom_linerange(aes(ymin = mean_l - margin_l, ymax = mean_l + margin_l), linewidth = 1) +
  geom_jitter(data = maples_2003, aes(x = watershed, y = stem_length), alpha = 0.3) +
  scale_color_manual(values = c("Reference" = "#E57B33", "Calcium-treated" = "#039199")) +
  labs(x = "Watershed", y = "Stem length (mm)") +
  theme_classic() +
  theme(legend.position = "none",
        text = element_text(family = "Times New Roman"),
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        plot.caption = element_text(hjust = 0),
        plot.caption.position = "plot",
        plot.title.position = "plot")
```

## f. Communicating about the results of a t-test

### figure caption

**Figure 1. Sugar maple stem lengths in calcium-treated and reference watersheds.** Stem lengths (mm) for calcium-treated (turquoise) and reference (orange) watersheds from Hubbard Brook Long-term Ecological Research site (HBR LTER). Dark points represent mean stem length and vertical lines represent confidence intervals with a 95% confidence level. Transparent points represent stem lengths.

### in-text references

There is a moderate (Cohen's d = 0.49) but significant effect of calcium treatment on sugar maple stem lengths (Student's t-test, t(238) = 3.78, _p_ < 0.001, $\alpha$ = 0.05). On average, sugar maple stem lengths in calcium-treated watersheds were 6.9 mm longer than those in reference watersheds (95% confidence interval: [3.3, 10.5] mm, Figure 1). 

### bonus in-text references, using in-line R

**When rendering your document, compare the text above with the text below. Are there any differences?**  

There is a moderate (Cohen's d = `r round(d_effsize$estimate, digits = 2)`) but significant effect of calcium treatment on sugar maple stem lengths (Student's t-test, t(`r length_ttest$parameter`) = `r round(length_ttest$statistic, digits = 1)`, _p_ < 0.001, $\alpha$ = 0.05). On average, sugar maple stem lengths in calcium-treated watersheds were `r round(mean_w1_2003 - mean_ref_2003, digits = 1)` mm longer than those in reference watersheds (CI = [`r round(length_ttest$conf.int, digits = 1)`] mm, Figure 1). 
